---
title: 'Cleaning up NOEP data'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ne-prep/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

## Summary

This script takes individual state-level .csv data that were manually downloaded from NOEP (http://www.oceaneconomics.org/Market/ocean/oceanEcon.asp) and combines them and cleans them for layer preparation. This data is used to create the **jobs**, **wage** and **gdp** layers for use in Livelihoods & Economies as well as Tourism & Recreation.

---

## Data Source

### National Ocean Economics Program (NOEP)

**Downloaded**: [Manually downloaded by state from website](http://www.oceaneconomics.org/Market/ocean/oceanEcon.asp) on July 3, 2018.    
**Description**:  Total number of jobs and wages per sector for RI, ME, MA, CT, and NH counties from 2005 to 2015. The data also include number of establishments and GDP for each sector - state - year.  
**Native data resolution**: County level     
**Time range**: 2005 - 2015  
**Format**:  Tabular  
---

## Setup

```{r, message = F, warning = F}
knitr::opts_chunk$set(fig.width = 10, fig.height = 6, fig.path = 'figs/', message = FALSE, warning = FALSE)

library(tidyverse)

#load common file
dir_git <- '~/github/ne-prep'
source(file.path(dir_git, 'src/R/common.R'))  ### an OHI-NE specific version of common.R

#set dir_anx to mazu
dir_anx <- file.path(dir_M, 'git-annex/neprep')
```

## Get data  

Load the data and fix the column classes to be numeric instead of character.
```{r}
## get a list of the files we're going to combine. These are all held on the Mazu server at NCEAS
files <- list.files(file.path(dir_anx, "_raw_data/NOEP/d2018"), full.names = T)

## read in each csv file and change columns from character to numeric while also removing commas and dollar signs ($)
out <- map_df(files, read_csv) %>%
  filter(!is.na(State)) %>% #there are some NA for state because these rows have notes
  mutate(Establishments = as.numeric(gsub(",", "", Establishments)),
         Employment     = as.numeric(gsub(",", "", Employment)),
         Wages          = gsub(",", "", Wages),
         GDP            = gsub(",", "", GDP)) %>%
  mutate(Wages          = as.numeric(gsub("[\\$,]", "", Wages))) %>% #doing a stepwise fix on gdp and wages for the $ and ,
  mutate(GDP            = as.numeric(gsub("[\\$,]", "", GDP))) 
```

Since we have two separate Massachusetts regions, we need to split the data according to how those counties border those two regions. The [`ma_ctys` table](https://github.com/OHI-Northeast/ne-prep/blob/gh-pages/src/tables/MA_counties.csv) lists what counties in Massachusetts belong to each region. For counties that are in both (Plymouth and Barnstable) we will simply divide the values by 2.


```{r split_MA}
# need to read in the MA counties since they are divided across two bioregions

ma_ctys <- read_csv("~/github/ne-prep/src/tables/MA_counties.csv") %>%
  select(-X1)

clean_data <- out %>%
        left_join(rgn_data, by = c("State" = "state")) %>%
        left_join(ma_ctys, by = c("rgn_name", "County", "rgn_id")) %>%
        mutate(Employment     = ifelse(State == "Massachusetts" & County %in% c("Plymouth", "Barnstable"), 
                                      Employment/2, Employment),
               Establishments = ifelse(State == "Massachusetts" & County %in% c("Plymouth", "Barnstable"), 
                                       Establishments/2, Establishments),
               Wages          = ifelse(State == "Massachusetts" & County %in% c("Plymouth", "Barnstable"), 
                                       Wages/2, Wages),
               GDP            = ifelse(State == "Massachusetts" & County %in% c("Plymouth", "Barnstable"),
                                       GDP/2, GDP))
```

## Save

The final output if `clean_noep_data.csv` in the data folder.

```{r save}
write.csv(clean_data, "data/clean_noep_data.csv")
```
